---
title: "Code comparison between Koschorreck et al. and Hall lab code for calculating CO~2~ concentration after headspace equilibration"
authors: "M. T. Trentman, J. R. Blaszczak, R. O. Hall, Jr."
date: "10/14/2020"
output:
  html_document: default
  pdf_document: default
---
Below we compare major differences between our group (Hall lab) and that of Koschorreck et al. for calculating the CO~2~ concentration in a water sample after headspace equilibration as a part of our comment to their manuscript in review with Biogeosciences. We compare the approaches used by each group and show that, generally speaking, the different approaches presented here result in the same outcome.  

## Calculating H^+^ and DIC at equilibrium
Both the Hall lab and Koschorreck et al. appraches use the total alkalinity equation to calculate the H^+^ concentration in the sample, but with different assumptions. The Hall lab assumes that H^+^ and OH^-^ are approximately zero, while Koschorreck et al. include them both in the total alkalinity equation. The assumption that H^+^ and OH^-^ are zero have little if any effect on the accuracy of the H^+^ estimation and allows for solving a 2nd order polynomial, which can be solved algebraically using the quadratic equation (full derivation can be found at https://github.com/jrblaszczak/CO2_headspace_code/tree/master/Code). Koschorreck et al. solve the much more complex 4th order polynomial using the polyroot function. Both approaches result in a similar H^+^ concentration.

The Hall lab and Koschorreck et al. use different calculations to estimate the DIC at equilibrium. Koschorreck et al. use the CO~2~ ionization (alpha) equation to calculate DIC, while the Hall lab uses the individual equations to calculate the carbonate, bicarbonate, and CO2 concentrations, and then adds them up to get DIC. Both approaches result in the same DIC concentration.

```{r warning=TRUE,error=TRUE}

###Raw code for each approach
##Kos
    h_all <- polyroot(c(-(2*K1*K2*co2),-(co2*K1+Kw),AT,1))
    real<-Re(h_all)
    h <-real[which(real>0)]
    DIC_eq <- co2 * (1 + K1/h + K1 * K2/(h * h)) 
    
##Hall
Carbfrom_C_A <- function(K1, K2, StmpCO2.umol , A){
  H <- (((-K1*StmpCO2.umol ))-sqrt(((K1*StmpCO2.umol )^2)-(4*-1*A*2*K1*K2*StmpCO2.umol )))/(2*-1*A)
  pH <- -1*log10(H)
  B <- (K1*StmpCO2.umol )/H
  Ca <- (K2*B)/H
  D <- StmpCO2.umol  + B + Ca
  A_check <- B + 2*Ca
  Carb1 <- list(H, pH, StmpCO2.umol , B, Ca, D, A, A_check)
  names(Carb1) <- c("H", "pH", "StmpCO2.umol ", "B", "Ca", "D", "A","A check")
  Carb1
  }

###Modified code to compare both DIC estimates using the same data
#Fake data
Kh<-0.03916
Kw<-7.1614*10^-15
K1<-4.1483*10^-7
K2<-4.199*10^-11
AT<-2000 #Kos label
A<-2000 #Hall label
co2<- 800 #Kos label
StmpCO2.umol<-800 #Hall label

#Run Kos
    h_all <- polyroot(c(-(2*K1*K2*co2),-(co2*K1+Kw),AT,1))
    real<-Re(h_all)
    h <-real[which(real>0)]
    DIC_eq <- co2 * (1 + K1/h + K1 * K2/(h * h)) 
    
#Run Hall lab
Carb1<-Carbfrom_C_A(K1, K2,StmpCO2.umol, A )

##Results
#Kos
h#1.66*10^-7
DIC_eq#2799

#Hall
Carb1$H#1.66*10^-07
Carb1$D#2799
```

## Calculating the original DIC 
Both the Hall lab and Koschorreck et al. have mostly similar approaches for estimating the original DIC of the water sample. The minor difference is that Koschorreck et al. do not include the density of water in their calculation (See Dickson et al. 2007 for more detail). This has no effect on the calculation of DIC in freshwater where the density is ~1.

```{r warning=TRUE,error=TRUE}
###Raw code for each approach
#Kos
DIC_ori <- DIC_eq + (pCO2_eq[i] - pCO2_headspace[i])/1000000/(R*(temp_eq[i]+273.15))*HS.ratio

#Hall
DIC_correction<-function(CO2_pre,CO2_post,vol_hs,temp_equil,vol_samp){
  delta_mol<-((CO2_post)-(CO2_pre)/1000000)*(vol_hs)/(R*(temp_equil + 273.15))
  delta_DIC<-delta_mol/(dens*vol_samp)
  delta_DIC
}
DIC_corr<-Carb1$D+delta_DIC

###Modified code to compare both DIC estimates using the same data
#Fake data
R <- 0.082057338
dens=0.9998395
pCO2_eq=500 #Kos label
CO2_post=500 #Hall label
pCO2_headspace=0 #Kos label
CO2_pre=0 #Hall label
temp_eq=20 #Kos label
HS.ratio=1.75
DIC_eq=2799
temp_equil=20 #Hall label
vol_samp=40
vol_hs=70

#Kos
DIC_corr=(pCO2_eq - pCO2_headspace)/1000000/(R*(temp_eq+273.15))*HS.ratio
DIC_corr##3.637481e-05

#Hall
DIC_correction<-function(CO2_pre,CO2_post,vol_hs,temp_equil,vol_samp){
  delta_DIC<-(((CO2_post-CO2_pre)/1000000))/(R*(temp_equil + 273.15))*(vol_hs/vol_samp)
  delta_DIC
  }
DIC_corr<-DIC_correction(CO2_pre,CO2_post,vol_hs,temp_equil,vol_samp)
DIC_corr# 3.637481e-05

```

## Calculating the original CO~2~ from the corrected DIC
Like the DIC equilibrium calculation above, the Hall lab and Koschorreck et al. use different calculations to calculate the original CO~2~ concentration.
```{r warning=TRUE,error=TRUE}
#Kos
h_all <- polyroot(c(-(K1*K2*Kw),K1*K2*AT-K1*Kw-2*DIC_ori*K1*K2,AT*K1-Kw+K1*K2-DIC_ori*K1,AT+K1,1))
    real<-Re(h_all)
    h <-real[which(real>0)]
    
    co2 <- h* (DIC_ori * h * K1/(h * h + K1 * h + K1 * K2)) / K1
   

#Hall
Carbfrom_D_A <- function(K1, K2, DIC_corr, A){
  a <- A
  b <- K1*(A-DIC_corr)
  c <- (A-(2*DIC_corr))*K1*K2
  H_t <- ((-1*b)+sqrt((b^2)-(4*a*c)))/(2*a)
  
  pH_t <- -1*log10(H_t)
  
  B_t <- (DIC_corr*K1*H_t)/((H_t^2)+(K1*H_t)+(K1*K2))
  Ca_t <- (DIC_corr*K1*K2)/((H_t^2)+(K1*H_t)+(K1*K2))
  
  C_t <- (H_t*B_t)/K1
  D2 <- C_t + B_t + Ca_t
  A2_check <- B_t + 2*Ca_t
  D_check <- C_t + B_t + Ca_t
  Carb2 <- list(H_t, pH_t, B_t, Ca_t, C_t, D_check, A2_check)
  names(Carb2) <- c("H", "pH", "B", "Ca", "CO2", "D_check", "A check")
  return(Carb2)
  }

##Test with fake data
#Other variables carried over from above
DIC_ori=3000 #Kos label
DIC_corr=3000 #Hall label
#Kos
h_all <- polyroot(c(-(K1*K2*Kw),K1*K2*AT-K1*Kw-2*DIC_ori*K1*K2,AT*K1-Kw+K1*K2-DIC_ori*K1,AT+K1,1))
    real<-Re(h_all)
    h <-real[which(real>0)]
    co2 <- h* (DIC_ori * h * K1/(h * h + K1 * h + K1 * K2)) / K1
co2 #1000.404

#Hall
Carb2<-Carbfrom_D_A(K1, K2, DIC_corr, A)
Carb2$CO2 ## 1000.404

```
