---
output: html_document
---

### Freshwater Carbonate Equilibration Equations
#### J.R. Blaszczak, R.O. Hall, M.T. Trentman
#### `r Sys.Date()`


Carbon dioxide is one multiple species of carbon that makes up the freshwater dissolved inorganic carbon (DIC) pool. The total DIC pool is the sum of carbon dioxide ($CO_2$), bicarbonate ($HCO_{3}^-$), and carbonate ($CO_{3}^{2-}$) concentrations.The strength of the carbonate buffering system is dependent on pH and alkalinity.

**There are a series of four equilibrium equations that describe the carbonate buffering system, as follows:**

(1) $K_{1} = \frac{[HCO_{3}^-][H^+]}{[H_{2}CO_{3}]}$


(2) $K_{2} = \frac{[CO_{3}^2-][H^+]}{[HCO_{3}^-]}$


(3) $A = [HCO_{3}^-] + 2[CO_{3}^2-]$


(4) $D = [H_{2}CO_{3}] + [HCO_{3}^-] + [CO_{3}^{2-}]$


Where $[H_{2}CO_{3}] \approx [CO_{2}]$ measured in the headspace in units $ppmv$, $A$ is streamwater alkalnity concentrations in units $mol$$m^{-3}$, $D$ is the total dissolved inorganic carbon concentrations, and the equilibrium constants $K_1$ and $K_2$ are temperature dependent and can be adjusted according to the following relationships (Millero et al. 2006):

$K_1 = 10^{ (-126.34048 + 6320.813/(T+273.15)+19.568224\times log(T+273.15)}$

$K_2 = 10^{ (-90.18333+5143.692/(T+273.15)+14.613358\times log(T+273.15))}$

Where $T$ is water temperature at equilibration in degrees Celsius.


<br/>

**First, we will use the measured alkalinity and $CO_{2}$ concentrations to solve for the first distribution of carbonate species.**


Start by rearranging $K_{1}$ to solve for $[HCO_{3}^-]$:

$[HCO_{3}^-] = \frac{K_{1}[CO_{2}]}{[H^+]}$

Next, rearrange $K_{2}$ to solve for $[CO_{3}^{2-}]$:

$[CO_{3}^{2-}] = \frac{K_{2}[HCO_{3}^-]}{[H^+]}$

Then, substitute $[HCO_{3}^-]$ in to get:

$[CO_{3}^{2-}] = \frac{K_{1}K_{2}[CO_{2}]}{[H^+]^2}$

To solve for $[H^+]$, plug $[HCO_{3}^-]$ and $[CO_{3}^{2-}]$ into the alkalinity ($A$) equation:

$A = \frac{K_{1}[CO_{2}]}{[H^+]} + 2\frac{K_{1}K_{2}[CO_{2}]}{[H^+]^2}$

Next, multiply the full equation by $[H^+]^2$ and arrange everything on the same side of the equation:

$0 = -A[H^+]^2 + K_{1}[CO_{2}][H^+] + 2K_{1}K_{2}[CO_{2}]$

Use the quadratic equation to solve for $[H^+]$:

$[H^+] = \frac{-K_{1}[CO_{2}] \pm \sqrt{(K_{1}[CO_{2}])^2 - 4(-A)(2K_{1}K_{2}[CO_{2}])}}{2\times(-A)}$

Use measured concentration values to calculate $[H^+]$, then use that value to calculate both $[HCO_{3}^-]$ and $[CO_{3}^{2-}]$ that can finally be plugged into the DIC equation:

$D = [CO_{2}] + [HCO_{3}^-] + [CO_{3}^2-]$

For good measure, double check whether your alkalinity concentration is recoverable by also plugging $[HCO_{3}^-]$ and $[CO_{3}^{2-}]$ into the alkalinity equation:

$A = [HCO_{3}^-] + 2[CO_{3}^2-]$

<br>

```{r eval=FALSE}
## In code, step 1 carbonate chemistry from C & A:
Carbfrom_C_A <- function(K1, K2, C, A){
  H_minus <- (((-K1*C))-sqrt(((K1*C)^2)-(4*-1*A*2*K1*K2*C)))/(2*-1*A)
  pH <- -1*log10(H_minus)
  B <- (K1*C)/H_minus
  Ca <- (K2*B)/H_minus
  D <- C + B + Ca
  A_check <- B + 2*Ca
  
  l <- list(H_minus, pH, C, B, Ca, D, A, A_check)
  names(l) <- c("H", "pH", "C", "B", "Ca", "D", "A","A check")
  return(l)
}

```


**Now that we have the first set of carbonate species to initialize the model, use the calculated DIC concentration from the previous step and the measured alkalinity concentration to run a forward model.**
Alkalinity is assumed to be constant throughout the 24 hour period.

First, let's solve for $[HCO_{3}^{-}]$ by rearranging the original $K_1$ and $K_2$ equations such that:

$[CO_2] = \frac{[HCO_{3}^{-}][H^+]}{K_1}$

and

$[CO_{3}^{2-}] = \frac{K_2[HCO_{3}^{-}]}{[H^+]}$

We can then plug both of these into the original $D$ equation:

$D = [CO_2] + [HCO_{3}^-] + [CO_{3}^{2-}]$

$D = \frac{[HCO_{3}^{-}][H^+]}{K_1} + [HCO_{3}^-] + \frac{K_2[HCO_{3}^{-}]}{[H^+]}$

Then solve the equation for $[HCO_{3}^-]$ and you get:

$[HCO_{3}^{-}] = \frac{K_{1}[H^+]D}{K_{1}[H^+] + K_{1}K_{2} + [H^+]^{2}}$

Repeat the same steps except solve for $[CO_{3}^{2-}]$ and you get:

$[CO_{3}^{2-}] = \frac{K_{1}K_{2}D}{[H^+]^{2} + K_{1}[H^+] + K_{1}K_{2}}$

In addition, solve for $[H^+]$ by plugging the above two solutions (clue: they have the same denominator so it is easier to solve) into the alkalinity equation:

$[H^+] = \frac{-K_{1}\times(A-DIC) + \sqrt{(K_{1}\times(A-DIC))^2 - 4(A)(K_{1}K_{2}*(A-2DIC))}}{2\times A}$

And fit the model to the measured $[CO_2]$ for the next time step using the equation:

$[CO_2] = \frac{[HCO_{3}^{-}][H^+]}{K_1}$


```{r eval=FALSE}
## In code, step 2 carbonate chemistry from D & A:
Carbfrom_D_A <- function(K1, K2, D1, A){
  a <- A
  b <- K1*(A-D1)
  c <- (A-(2*D1))*K1*K2
  H_t <- ((-1*b)+sqrt((b^2)-(4*a*c)))/(2*a)
  
  pH_t <- -1*log10(H_t)
  
  B_t <- (D1*K1*H_t)/((H_t^2)+(K1*H_t)+(K1*K2))
  Ca_t <- (D1*K1*K2)/((H_t^2)+(K1*H_t)+(K1*K2))
  
  C_t <- (H_t*B_t)/K1
  D2 <- C_t + B_t + Ca_t
  A2_check <- B_t + 2*Ca_t
  
  l <- list(H_t, pH_t, C_t, B_t, Ca_t, D2, A, A2_check)
  names(l) <- c("H", "pH", "C", "B", "Ca", "D", "A", "A check")
  return(l)
}
```




#### References

Millero et al. 2006









